sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-09 doc/paper-jss/paper-jss.Rmd
Start adding eqns

We begin with a simple species distribution model of encounter probability of
Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey conducts
depth-stratified random sampling biennially, resulting in a georeferenced
dataset. The dataset includes columns for year, latitude, longitude, Universal
Transverse Mercator (UTM) zone 9 coordinates, bottom depth, and encounter vs.
non-encounter.

@@

```{r pcod-head, echo=TRUE}
select(pcod, year, lat, lon, X, Y, depth, present) |> 
  head(n = 3)
```

An \pkg{sdmTMB} model requires a data frame that contains a response column,
columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant
projection such as the Universal Transverse Mercator (UTM) to ensure that
distance remains constant throughout the study region [e.g., using
`sf::st_transform()`, @pebesma2018]. Here we use the helper function
`add_utm_columns()` to add UTM coordinates with km units (so our estimated
spatial range parameter is not too big or small). By default, the function will
guess the UTM zone and create new columns `X` and `Y`. Since our example
data set already has these UTM columns, we can skip running this code.

@@

We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff`
is passed through to it. Soon, \pkg{sdmTMB} will switch to using the
\pkg{fmesher} package for this purpose, which is hosted on CRAN. Alternatively,
we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and
supplied it to the `mesh` argument in `make_mesh()`. We can inspect our mesh
object with the associated plotting method.

@@

Our random field model can be written as TODO.

@@

@@

TODO EQUATIONS

@@

We find that the Poisson-link-lognormal model is favoured with AIC. We next
test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist.

@@

@@

>>>>

We begin with a simple species distribution model of encounter probability of
Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey is conducted by
Fisheries and Oceans Canada (REF) and follows a depth-stratified random
sampling design, resulting in a georeferenced dataset. The data frame `pcod` is
available as package data in \pkg{sdmTMB}. Relevant columns include latitude,
longitude, Universal Transverse Mercator (UTM) coordinates, bottom
depth, and encounter (`present = 1`) vs. non-encounter (`present = 0`).

@@

```{r pcod-head, echo=TRUE}
select(pcod, lat, lon, X, Y, depth, present) |> 
  head(n = 3)
```

An \pkg{sdmTMB} model requires a data frame that contains a response column,
columns for any predictors, and columns for spatial coordinates. Usually it
makes sense to convert the spatial coordinates to an equidistant projection
such as UTMs to ensure that distance remains constant throughout the study
region [e.g., using `sf::st_transform()`, @pebesma2018]. Here we use the helper
function `add_utm_columns()` to add UTM coordinates with km units (so our
estimated spatial range parameter is not too big or small). By default, the
function will guess the UTM zone and create new columns `X` and `Y`. Since our
example data already has these UTM columns, we can skip running this code.

@@

\sloppy
We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` or
`fmesher::fm_rcdt_2d_inla()` and `cutoff` is passed through to these functions.
Alternatively, we could have created any mesh via \proglang{R}-\pkg{INLA} or
\pkg{fmesher} and supplied it to the `mesh` argument in `make_mesh()`. We can
inspect our mesh object with the associated plotting method.

@@

Our random field model can be written as

$$
\begin{aligned}
y_{\boldsymbol{s}} &\sim \operatorname{Bernoulli} \left(\mu_{\boldsymbol{s}}\right),\\
\mu_{\boldsymbol{s}} &= \operatorname{logit}^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s}} \boldsymbol{\beta} +
\omega_{\boldsymbol{s}}
\right),
\end{aligned}
$$

where TODO.

$$
\omega_{\boldsymbol{s}} \sim \mathrm{MVN}(\boldsymbol{0},\boldsymbol{\Sigma}_\omega)
$$

@@

$$
\begin{aligned}
\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
O_{\boldsymbol{s},t} +
\boldsymbol{X}^{\mathrm{tvc}}_{\boldsymbol{s},t} \gamma_t +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} 
\right),
\end{aligned}
$$

where TODO.

and TODO

$$
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right).
\end{aligned}
$$

@@

TODO EQUATIONS NEED TO INTRO THE POISSON LINK AS BRIEFLY AS POSSIBLE. SEE sdmTMB VIGNETTE.

@@

We find that the Poisson-link delta-lognormal model is favoured with AIC. We
next test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist from year to year.

@@

$$
\begin{aligned}
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}},  \:
\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where TODO.

@@

$$
\begin{aligned}
\mathbb{E}[y_{\boldsymbol{s},t}] &= \mu_{\boldsymbol{s},t},\\
\mu_{\boldsymbol{s},t} &=
f^{-1} \left( \boldsymbol{X}^{\mathrm{main}}_{\boldsymbol{s},t} \boldsymbol{\beta} +
\alpha_g +
\boldsymbol{X}^{\mathrm{svc}}_{\boldsymbol{s},t} \zeta_{\boldsymbol{s}} +
\omega_{\boldsymbol{s}} +
\epsilon_{\boldsymbol{s},t} \right),
\end{aligned}
$$

where TODO.
